Frustratingly easy multi-domain, multilingual translation with latent multi-task group dropout

### 2021-11-05 2022/acl/acl_latex.tex
add some insignificant details

\newcommand{\fyTodo}[1]{\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyTodostar}[1]{\Todo*[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyDone}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyFuture}[1]{\done[FY]\Todo[FY:]{\textcolor{red}{#1}}}
\newcommand{\fyDonestar}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\jcTodo}[1]{\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\jcDone}[1]{\done[JC]\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\revision}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondone}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondel}[1]{}
\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{{#1}}}

@@

We incorporate the latent variables $m_{l\in[0,\cdots,L]}^d$ into the log-likelihood of our model as below. For the sake of brevity, we denote $m_l^d$ the set of latent variables $m_{l\in[0,\cdots,L]}^d$
\begin{equation}
\hspace{-1.5em}
\begin{array}{rcl}
\text{log} P(y|x,\theta,d) & \text{=} & \text{log} P(y, m_l^d |x,\theta,d) \text{-} \text{log} P(m_l^d|y, x,\theta,d) \\
& \text{=} & \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y, m_l^d | x,\theta,d) \text{-} \text{log} P(m_l^d | \Phi_l^d) \big] \\
& \text{+} & KL(P(m_l^d | \Phi_l^d) \parallel P(m_l^d|y, x,\theta,d)) \\
& \geq & \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y, m_l^d | x,\theta,d) \text{-} \text{log} P(m_l^d | \Phi_l^d) \big]  \\
& \text{=} & \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y | x,\theta,m_l^d, d) \big] \\
& \text{-} & KL \big( P(m_l^d | \Phi_l^d) \parallel P(m_l^d | x,\theta,d) \big) \\
& \text{=} & \mathcal{L}(\theta, \Phi_l^d, x , y) \label{eq:lower-bound}
\end{array}
\end{equation}
our training objective will be the lowerbound, which is called the evidence lower bound (ELBO) in variational statistic \citep{Diederick14auto}, $\mathcal{L}(\theta, \Phi_l^d, x , y)$. The variational distribution of $m_l^d$, $P(m_l^d|\Phi_l^d)$, is defined as follows

@@

\begin{align}
\hspace{-2.em}
T_1 &= \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y | x,\theta,m_l^d, d)  \big] \nonumber \\
&= \mathbb{E}_{g_l^d(p) \displaystyle{\mathop{\sim}^{\text{i.i.d}}} Gumbel(0,1) \ } \big[ \text{log} P(y | x,\theta,\tilde{m}_l^d,d) \big] \nonumber
\end{align}
where
\begin{align}
\hspace{-2.em}
g_l^d & \in \mathbb{R}^{n_p} \nonumber \\
g_l^d(p) & \displaystyle{\mathop{\sim}^{\text{i.i.d}}} Gumbel(0,1) \nonumber \\
& \forall p \in \{ 0,\cdots,n_p\text{-}1 \} \nonumber \\
i_1, \cdots, i_k &= Top\_k \ \{ \ g_l^d(0) + \Phi_l^d(0), \cdots, \label{eq:top-k} \\ 
&g_l^d(n_p\text{-}1) + \Phi_l^d(n_p\text{-}1) \ \} \nonumber \\
\tilde{m}_l^d(p) &= \mathbb{I}(p \in Ind^d). \nonumber 
\end{align}
%
\begin{align}
T_2 &= - KL \big( P(m_l^d | \Phi_l^d) \parallel P(m_l^d | x,\theta,d) \big) \nonumber \\ 
	&= \mathbb{H} \big[ P(m_l^d | \Phi_l^d) \big] - \text{log}(\#\Delta^{n_p}_k) \nonumber \\ 
	&= \mathbb{H} \big[ P(i_1,\cdots,i_k | \Phi_l^d) \big] - \text{log}(\#\Delta^{n_p}_k) \nonumber \\ \
	&\geqslant \mathbb{H} \big[ P(i_1 | \Phi_l^d) \big] - \text{log}(\#\Delta^{n_p}_k) \label{eq:entropy}
\end{align}

@@

In Appendix \ref{appendix:a}, we will prove that $\lim_{\tau \rightarrow 0}\hat{m}_l^d(\tau) = \tilde{m}_l^d$. Furthermore, we will also provide the computation of $\hat{m}_l^d(\tau)$ and prove that $\hat{m}_l^d(\tau)$ is a differentiable function w.r.t $\Phi_l^d$ and its differentiation is computed via implicit differentiation theorem. During the training, we approximate $\tilde{m}_l^d$ by $\hat{m}_l^d(\tau)$ by gradually decaying the hyperparameter $\tau$ to $0$. The gradient of the term $T_1$ w.r.t $\Phi_l^d$ is computed via the chain rule as follows

@@

\frac{\partial T_1}{\Phi_l^d} = \frac{\partial T_1}{\hat{m}_l^d(\tau)} \times \frac{\hat{m}_l^d(\tau)}{\Phi_l^d}

@@

The gradient $\frac{\partial T_1}{\hat{m}_l^d(\tau)}$ is computed via autograd algorithm while $\frac{\hat{m}_l^d(\tau)}{\Phi_l^d}$ is computed via implicit differentiation, that we will explain carefully in Appendix \ref{appendix:a}.

@@

In both cases, we examinate the performance of three following MT systems.

@@

We reuse the same data of a recent work of \citet{Pham21revisiting} in multi-domain translation. For the sake of brevity, we leave most of the details of data in Appendix \ref{appendix:c}.

@@

According to Table \ref{tab:multilingual}, \system{LaMGD} shows a boost of 0.42, 0.33, 0.32 in average over the multilingual \system{Transformer} in O2M-related, M2O-related, M2O-diverse respectively. Significant gains are observed in \domain{bel}, \domain{glg} (both direction), \domain{hin} and \domain{bos} (O2M direction).

@@

    \system{LaMGD}  \hfill{\footnotesize[+0m]}  & 5.20&9.40&20.60&22.80&39.60&19.60&22.40&15.00&19.33 \\ 

@@

    \system{LaMGD}   \hfill{\footnotesize[+0m]}  &11.30&17.40&28.60&28.70&	43.70&24.50&30.70&25.60&26.31 \\ 

@@

    \system{LaMGD}   \hfill{\footnotesize[+0m]} &23.50&9.60&21.50&32.20&37.70&38.60&40.00&18.90&27.75 \\

>>>>

\newcommand{\fyTodo}[1]{\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyTodostar}[1]{\Todo*[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyDone}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\fyFuture}[1]{\done[FY]\Todo[FY:]{\textcolor{red}{#1}}}
\newcommand{\fyDonestar}[1]{\done[FY]\Todo[FY:]{\textcolor{orange}{#1}}}
\newcommand{\jcTodo}[1]{\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\jcDone}[1]{\done[JC]\Todo[JC:]{\textcolor{red}{#1}}}
\newcommand{\revision}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondone}[1]{\textcolor{black}{#1}}
\newcommand{\revisiondel}[1]{}
\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{{#1}}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}

@@

We incorporate the latent variables $m_{l\in[0,\cdots,L]}^d$ into the log-likelihood of our model as below. For the sake of brevity, we denote $m_l^d$ the set of latent variables $m_{l\in[0,\cdots,L]}^d$
\begin{align}
\hspace{-1.em}
\text{log} P(y|x,\theta,d) & \text{=} \text{log} P(y, m_l^d |x,\theta,d) \text{-} \text{log} P(m_l^d|y, x,\theta,d) \nonumber \\
& \text{=} \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y, m_l^d | x,\theta,d) \text{-} \text{log} P(m_l^d | \Phi_l^d) \big] \nonumber \\
& \text{+} KL(P(m_l^d | \Phi_l^d) \parallel P(m_l^d|y, x,\theta,d)) \nonumber \\
& \geqslant \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y, m_l^d | x,\theta,d) \text{-} \text{log} P(m_l^d | \Phi_l^d) \big] \nonumber \\
& \text{=} \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y | x,\theta,m_l^d, d) \big] \label{eq:lower-bound} \\
& \text{-} KL \big( P(m_l^d | \Phi_l^d) \parallel P(m_l^d | x,\theta,d) \big) \nonumber \\
& \text{=} \mathcal{L}_{\text{ELBO}}(\theta, \Phi_l^d, x , y). \nonumber
\end{align}
$\mathcal{L}_{\text{ELBO}}(\theta, \Phi_l^d, x , y)$ is the lowerbound, which is called the evidence lower bound (ELBO) in variational statistic \citep{Diederick14auto}, of our log-likelihood. Our training loss will be $-\mathcal{L}_{\text{ELBO}}(\theta,\Phi_l^d,x,y)$.

The variational distribution of $m_l^d$, $P(m_l^d|\Phi_l^d)$, is defined as follows

@@

\begin{align}
\hspace{-2.em}
T_1 &= \mathbb{E}_{P(m_l^d | \Phi_l^d)} \big[ \text{log} P(y | x,\theta,m_l^d, d)  \big] \nonumber \\
&= \mathbb{E}_{g_l^d(p) \displaystyle{\mathop{\sim}^{\text{i.i.d}}} Gumbel(0,1) \ } \big[ \text{log} P(y | x,\theta,\tilde{m}_l^d,d) \big] \nonumber
\end{align}
where
\begin{align}
\hspace{-2.em}
g_l^d & \in \mathbb{R}^{n_p} \nonumber \\
g_l^d(p) & \displaystyle{\mathop{\sim}^{\text{i.i.d}}} Gumbel(0,1) \nonumber \\
& \forall p \in \{ 0,\cdots,n_p\text{-}1 \} \nonumber \\
i_1, \cdots, i_k &= Top\_k \ \{ \ g_l^d(0) + \Phi_l^d(0), \cdots, \label{eq:top-k} \\ 
&g_l^d(n_p\text{-}1) + \Phi_l^d(n_p\text{-}1) \ \} \nonumber \\
\tilde{m}_l^d(p) &= \mathbb{I}(p \in Ind^d). \nonumber 
\end{align}
%
\begin{align}
\hspace{-2.em}
T_2 &= - KL \big( P(m_l^d | \Phi_l^d) \parallel P(m_l^d | x,\theta,d) \big) \nonumber \\ 
	&= \mathbb{H} \big[ P(m_l^d | \Phi_l^d) \big] - \text{log}(\#\Delta^{n_p}_k) \nonumber \\ 
	&= \mathbb{H} \big[ P(i_1,\cdots,i_k | \Phi_l^d) \big] - \text{log}(\#\Delta^{n_p}_k) \nonumber \\ \
	&\geqslant \mathbb{H} \big[ P(i_1 | \Phi_l^d) \big] - \text{log}(\#\Delta^{n_p}_k) \label{eq:entropy}
\end{align}

@@

In Appendix \ref{appendix:a}, we will prove that $\lim_{\tau \rightarrow 0}\hat{m}_l^d(\tau) = \tilde{m}_l^d$. Furthermore, we will also provide the computation of $\hat{m}_l^d(\tau)$ and prove that $\hat{m}_l^d(\tau)$ is a differentiable function w.r.t $\Phi_l^d$ and its differentiation is computed via implicit differentiation theorem. During the training, we approximate $\tilde{m}_l^d$ by $\hat{m}_l^d(\tau)$ by gradually decaying the hyperparameter $\tau$ to $0$. The gradient of the term $T_1$ w.r.t $\Phi_l^d$ is computed via the chain rule as follows
\begin{equation}
\frac{\partial T_1}{\Phi_l^d} = \frac{\partial T_1}{\hat{m}_l^d(\tau)} \times \frac{\hat{m}_l^d(\tau)}{\Phi_l^d}
\end{equation}
The gradient $\frac{\partial T_1}{\hat{m}_l^d(\tau)}$ is computed via autograd algorithm while $\frac{\hat{m}_l^d(\tau)}{\Phi_l^d}$ is computed via implicit differentiation, that we will explain carefully in Appendix \ref{appendix:a}.

We jointly train the Transformer parameters $\theta$ and the parameters of the latent dropping masks $\Phi_l^d$ using the following multi-task loss.

@@

\mathcal{L}(\theta,\Phi_l^1,\cdots,\Phi_l^k) = \displaystyle{\mathop{\sum}_{d=1}^{k}} \mathbb{E}_{x \sim \mathcal{D}_s^d,y \sim g^d(x)} \big[-\mathcal{L}_{\text{ELBO}}(\theta,\Phi_l^d,x,y)\big]

@@

in which $\mathcal{D}_s^d$ is distribution of the task $d$ over the input space $\Omega^d_s$; $g^d: \Omega^d_s \rightarrow \Omega^d_t$ is the mapping function, which our multi-task modern needs to learn, of the task $d$; $-\mathcal{L}_{\text{ELBO}}(\theta,\Phi_l^d,x,y)$ is the ELBO loss, which we defined in Equation \ref{eq:lower-bound}.

@@

In both cases, we examine the performance of three following MT systems.

@@

We reuse the same data as the recent work of \citet{Pham21revisiting} in multi-domain translation. For the sake of brevity, we leave most of the details of data in Appendix \ref{appendix:c}.

@@

According to Table \ref{tab:multilingual}, \system{LaMGD} shows a boost of 0.42, 0.33, ,0.32 in average over the multilingual \system{Transformer} in O2M-related, M2O-related, M2O-diverse respectively. Significant gains are observed in \domain{bel}, \domain{glg} (both direction), \domain{hin} and \domain{bos} (O2M direction). However, \system{LaMGD} lags behind multilingual \system{Transformer} in O2M-diverse. 

@@

    \system{LaMGD}  \hfill{\footnotesize[+0m]}  & 5.20&\SB{9.40}&20.60&\SB{22.80}&39.60&19.60&22.40&15.00&19.33 \\ 

@@

    \system{LaMGD}   \hfill{\footnotesize[+0m]}  &11.30&\SB{17.40}&28.60&\SB{28.70}&43.70&24.50&30.70&25.60&26.31 \\ 

@@

    \system{LaMGD}   \hfill{\footnotesize[+0m]} &\SB{23.50}&9.60&\SB{21.50}&32.20&37.70&38.60&40.00&18.90&27.75 \\
